### 2020-12-04 doc/manuscript/Dow_manuscript.Rmd
replace word table with .png, insert tree-ring methods, misc

\newpage
*(Do we want a table of species here?? Columns would be Species, xylem porosity, n trees, n tree-years (after cleaning). Details will be in an SI table)*

```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
library(knitr)
species_table <- read.csv("tables_figures/species_table.csv",stringsAsFactors = FALSE, check.names = FALSE)
knitr::kable(species_table) 
```

@@

(*The following seems out of date, right? Scripted cleaning is not described.* -- *Manually cleaning was done for pre-analysis data. Added some text to clarify.*)
The dendroband data were manually cleaned by visual inspection before being used for modeling. Three classes of mistakes: 1. Error in measurement - weekly measurement was drastically different from previous week and following week. Mistakes were removed from the raw data before modeling. 2. Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern. Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year. In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level. 3. Other - cases where data was clearly wrong but with unknown causes. In cases where several measurements were an issue, they were simply removed. If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.

We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig. `r schematic`). This 5 parameter logistic model was fitted to each tag year's dendroband measurements in R: L + ((K - L) / (1 + 1/theta * exp(-(r * (doy - doy_ip) / theta)) ^ theta))
Where L and K are lower and upper asymptotes of the model, theta is a tuning parameter controlling the slope of the curve toward the upper asymptote, r is the slope of the curve at the inflection point, doy_ip is the day of year where the inflection point occurs, and doy are the modeled days of year, in this case, 1-365. *We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow.* Using the optimized parameters for each tree year, we could then determine the DBH on every day within that year, giving us the ability to extract the intra-annual growth variables we were interested in.

@@

### Tree-rings
(*tree rings-- KAT can write this*)

@@

## ("limitation in generalization, how to interpret/fix")
THe tree-rings also reveal that the strongest temperature sensitivity aligned with peak growing season.
This pattern is common across tree-ring studies in temperate deciduous forests.
(*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree grwoth will hold generall.y.

@@

We have shown that intraannual growth patterns in North American temperate trees change in response to increasing spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likeliehood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994). The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014). The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994). This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.

>>>>

```{r species_table, eval=TRUE, echo=FALSE, warning=FALSE}
knitr::include_graphics("tables_figures/species_table.png")
```

@@

The dendroband data were manually cleaned by visual inspection before analysis. Three classes of mistakes: 1. Error in measurement - weekly measurement was drastically different from previous week and following week. Mistakes were removed from the raw data before modeling. 2. Band slip or stuck - measurements freeze and remain unchanged until sudden jump followed by normal growth pattern. Band slips were sometimes followed by measurements indicating the band became stuck; in these cases, the tree was thrown out for the corresponding year. In cases where the slip was followed by normal growth, the initial slip-point was removed, and the following points were shifted down to the pre-slip level. 3. Other - cases where data was clearly wrong but with unknown causes. In cases where several measurements were an issue, they were simply removed. If there were no clear solutions to ‘fix’ the data, the entire year was removed from the analysis.

We fit the model of @mcmahon_general_2015 to define *phenological dates and growth rates* (Fig. `r schematic`). This 5 parameter logistic model was fitted to each tag year's dendroband measurements in R: 
$$L + ((K - L) / (1 + 1/\theta * exp(-(r * (doy - doy_{ip}) / \theta)) ^ {\theta}))$$  **(Cameron, please check formatting. It should be possible to write this more clearly with latex formatting. )**
Here, $L$ and $K$ are lower and upper asymptotes of the model, theta is a tuning parameter controlling the slope of the curve toward the upper asymptote, $r$ is the slope of the curve at the inflection point, $doy_{ip}$ is the day of year where the inflection point occurs, and $doy$ are the modeled days of year, in this case, 1-365. *We note that the maximum growth rate does not always occur on the doy_ip, and is controlled magically by theta somehow.* Using the optimized parameters for each tree year, we could then determine the DBH on every day within that year, giving us the ability to extract the intra-annual growth variables we were interested in.

@@

### Tree-rings
Dominant tree species were cored at both sites following sampling designs that covered a broad range of $DBH$.
We analyzed records for the dominant ring- and diffuse- porous species at each site (Table `r species_table`), excluding species with other xylem architectures (*Juglans nigra* L. at SCBI, *Tsuga Canadensis* (L.) Carrière at Harvard Forest)
At SCBI,  were collected as described in @helcoski_growing_2019. 
Briefly, ...  (*summarize / reword from Helcoski*). 
In total, we analyzed # cores representing eleven of the twelve species contributing most to aboveground woody productivity within the SCBI ForestGEO plot.
At Harvard Forest, ...(*Neil can write this easily, or KAT probably has the info*)
In total, we analyzed # cores from three species.

All cores were mounted, sanded, cross-dated, and measured using standard methodology. 
(*describe chronology building-- as in @helcoski_growing_2019-- Neil can do this quickly*)
Start year was defined at SSS=0.8.

@@

## ("limitation in generalization, how to interpret/fix")
THe tree-rings also reveal that the strongest temperature sensitivity aligned with peak growing season.
This pattern is common across tree-ring studies in temperate deciduous forests.
(*1-2 sentences generalizing about climate sensitivity of temperate deciduous forest growth--Neil can do this most easily.*)
Thus, while our analysis is limited to two sites, we expect that the finding that spring temperatures have limited effect on annual tree growth will hold generally.

@@

We have shown that intraannual growth patterns in North American temperate trees change in response to increasing spring temperatures. The magnitude of these changes depends largely on the evolutionary differences inherent in the two wood types analyzed here. Ring-porous trees must repair winter embolism of their large xylem vessels before leaf expansion, achieving roughly 25% total yearly growth simultaneously with bud burst, and 50% yearly growth around the time of 75% leaf emergence (D'Orangeville et al,). On the other hand, diffuse-porous trees are less prone to cavitation and so do not need to repair winter embolism before bud burst, allowing up to 75% leaf out before reaching their 25% yearly growth totals (D'Orangeville et al). These differences are likely evolutionary hydraulic adaptations to climate (Baas & Wheeler, 2011). The presence of wide-xylem cells in ring-porous trees allows for greater water conductivity in the spring at the cost of increased cavitation risk during spring drought, and high likelihood of damage during frost events (Cochard & Tyree, 1990; Sperry et al, 1994). The pre-leaf stem growth occurring in these species is fueled by non-structural carbon stores, often used for other cell maintenance activities within the plant (Sperry et al, 1994; Palacio et al, 2014; Sevanto et al, 2014). The dependence on these carbon stores for cambial activation could prove to be detrimental to long-term health of the tree, as exhaustion of reserves leads to carbon starvation (Sevanto et al, 2014). Diffuse-porous trees have narrower xylem cells lowering the risk of cavitation, but leading to a decreased water conductivity in the spring, revealing a growth strategy more focused on a relatively faster, shorter growing season in the Summer (Sperry et al, 1994). This allows the diffuse-porous trees to conserve their carbon stores, making them potentially more resistant to pest/pathogen invasions?, but exposing them to higher temperatures and drought risk during their Summer-centered growth period.
